Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-28 rnni_convexity.tex
update subsection mdtree

\todo{Lena: Think if we want to change the following two paragraphs to when we first pull $x_n$ up, then $x_{n-1}$, etc.}
The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively such that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
Specifically, in step $k$ ($k = 0, \ldots, n-3$) taxon $x_{n-k}$ is moved up to the position it has in $R$.
Here we say that a taxon moves up when it swaps position with its neighboured taxon, which corresponds to an $\nni$ move.
Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
After $n-2$ steps we reach $R$.

The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
\todo{Lena: If you decide to change the algorithm, mention here that what we've implemented is different from what is described here.}
\todo{Alex: I changed the algorithm. It shouldn't be too much work to change the implementation and the pseudo-code here. Do you want to keep the old version or shall I change everything to the new version?}
\end{description}}
Note that any algorithm with this property returns a shortest caterpillar path.

@@

The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
It is called $\mdtree$, which stands for Maximum Distance Tree, as it computes a tree with maximum distance if the input tree is a caterpillar tree.
This will be discussed in Section~\ref{section:diameter}.

$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows:
At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
Though this representation appears similar to the list representation of caterpillar trees, it is ambiguous since we do not specify how to order cherry taxa.
$\mdtree$ constructs an output tree $R$ as follows:
Initially, $R$ only consists two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.

@@

Both the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined within $\mdtree$.
Therefore, this algorithm is non-deterministic which is an important observation needed in Section~\ref{section:diameter} where this algorithm is used for finding the radius of $\rnni$.

>>>>

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively such that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
Specifically, in step $k$ ($k = 0, \ldots, n-3$) taxon $x_{n-k}$ is moved up to the position it has in $R$.
Here we say that a taxon moves up when it swaps position with its neighboured taxon, which corresponds to an $\nni$ move.
Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
After $n-2$ steps we reach $R$.

The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
\todo{Lena: If you decide to change the algorithm, mention here that what we've implemented is different from what is described here.}
\todo{Alex: I changed the algorithm. It shouldn't be too much work to change the implementation and the pseudo-code here. Do you want to keep the old version or shall I change everything to the new version?}
Note that any algorithm with this property returns a shortest caterpillar path.

@@

The algorithm $\mdtree$ (Maximum Distance Tree) will be an important tool for finding the radius of the $\rnni$ graph.
It computes a tree with maximum distance if the input tree is a caterpillar tree, which we will prove in Section~\ref{section:diameter}.

$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows:
At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
This list does uniquely identify trees since we do not specify how to order taxa that share a parent.
$\mdtree$ constructs an output tree $R$ as follows:
Initially, $R$ only consists two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.

@@

Both the list computed in Line~\ref{alg:mdtree_lineL} and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined within $\mdtree$.
Therefore, this algorithm is non-deterministic, which is an important observation that we will use for finding the radius of $\rnni$ in Section~\ref{section:diameter}.
